---
title: "MAGIC AI Workshop 03 Natural Language Processing"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(888)
```

## Introduction

In this lab, we will introduce tools for natural language processing (NLP), from basic data preparation through to some exploration and building a simple machine learning model. We are only scratching the surface of what is possible with NLP methods in this lab. See the tidytext website for further examples. 

You'll need several packages for the lab including:

- `tidytext`: a library for cleaning and processing text data
- ?? SOME OTHER ONES

```{r eval=FALSE}
install.packages(c("tidytext"))
```

Now load the first few packages:

```{r message=FALSE, warning=FALSE}
library(tidytext)
```

## Data

We'll use a set of tweets related to climate change from 2015 to 2018, taken from:

https://www.kaggle.com/datasets/edqian/twitter-climate-change-sentiment-dataset

The data are held in the file *twitter_sentiment_data.csv*, which you can download from the github repository. Read these in and take a quick look. There are three columns: a sentiment estimate, the tweet (`message`) and a tweet id. The sentiment estimate was provided by a group of experts and are tagged as follows:

- `2` (News): the tweet links to factual news about climate change
- `1` (Pro): the tweet supports the belief of man-made climate change
- `0` (Neutral): the tweet neither supports nor refutes the belief of man-made climate change
- `-1`(Anti): the tweet does not believe in man-made climate change

```{r}
dat <- read.csv("../datafiles/twitter_sentiment_data.csv")
```

